Review and comparison of tools for manageing X.509 type keys and certificates

### 2016-11-25 TEX/draft.tex
lists in overview, finalized Conclusion with best tools mentioned.

\section{General}
\begin{table}[h!]
\centering
\caption{Overwiew General}
\label{my-label}
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
                              & \multicolumn{5}{c|}{General}                                                                                   \\ \cline{2-6} 
\multirow{-2}{*}{Tools}       & Type   & View information about certificate                 & License & Operating System                        & Tool Type           \\ \hline
OpenSSL                       & PKCS12 & \cellcolor[HTML]{34FF34}Yes                        & Public  & \cellcolor[HTML]{34FF34}Both            & Command Line        \\ \hline
Keytool                       & JKS    & \cellcolor[HTML]{34FF34}{\color[HTML]{000000} Yes} & Public  & \cellcolor[HTML]{34FF34}Both            & Command Line        \\ \hline
GnuPG                         & None   & \cellcolor[HTML]{FE0000}No                         & Public  & \cellcolor[HTML]{34FF34}Both            & Command Line        \\ \hline
Keystore Explorer             & Both   & \cellcolor[HTML]{34FF34}Yes                        & Public  & \cellcolor[HTML]{FFC702}Win $\vert$ OSX & Graphical Interface \\ \hline
XCA                           & PKCS12 & \cellcolor[HTML]{34FF34}Yes                        & Public  & \cellcolor[HTML]{34FF34}Both            & Graphical Interface \\ \hline
GnuTLS                        & PKCS12 & \cellcolor[HTML]{34FF34}Yes                        & Public  & \cellcolor[HTML]{FFC702}UNIX            & Command Line        \\ \hline
Makecert                      & PKCS12 & \cellcolor[HTML]{FE0000}No			    & Public  & \cellcolor[HTML]{FFC702}Windows         & Command Line        \\ \hline
Windows PowerShell PKI Module & PKCS12 & \cellcolor[HTML]{34FF34}Yes			    & Public  & \cellcolor[HTML]{FFC702}Windows         & Command Line        \\ \hline
\end{tabular}
\end{adjustbox}
\end{table}
%TABLE END

\subsection{Type}
This section defines type of storage file in which the certificates or keys are saved. Most common are PKCS\#12 (.pfx and .p12 extentions) and JKS (Java KeyStore).

\subsection{View information about certificate}
This section shows whether it is possible to view information certificates and keys and additional information. 
    \begin{itemize}
     \item \textit{Yes} - viewing information about certificates and keys is availiable. This includes length, used algorithm, fingerprint.
     \item \textit{No}  - tool does not have an option to view any information about certificates or keys.
    \end{itemize}
\subsection{License}
Type of license and possibility of using said tool for testing or production code. Public means that license is not requiered for use in production code but does not mean that it should be used as such. Definition if it is advised to use said tool in production code or only in testing enviroment will be talked about in next chapters under each tool.

\subsection{Operating Systems}
In this part the Operating System where the tool is imported is defined. Both means that it is implemented under UNIX like systems and Windows systems. Otherwise the Operating System is defined (Windows $\vert$ UNIX).

@@

\paragraph{Self-signed certificate}
The tool is capable of generating private key and corresponding self-signed certificate

\paragraph{Certificate Signing Request}
Possibility of using a command to generate key pair and certificate signing request to certificate authority.

\paragraph{Specify length}
Possibility to specify the length of output key

\paragraph{Specify algorithm}
Possibility to choose between different types of algorithms for key generation and message digest\textit{(hash)}.

@@

\paragraph{Specify Type}
Specify if generated certificate will belong to certificate authority or whether it will be end certificate

\paragraph{Specify Path Length}
Specify the maximum length of certificate authority chain

@@

\paragraph{Certificate Signing Request signing}
Possibility of signing a certificate signing request with certificate authority's key

\paragraph{Create combination of private key and signed chain}
Possibility of generating private key and chain signed by certificate authority that will be outputed to a single file

\paragraph{Specify certificate validity}
{ossibility of choosing how long the certificate will be valid. This should be done by certificate authority.

\paragraph{Setting Subject Alternative Name for end certificates}
Possibility of choosing Subject Alternative Name for end certificates. That should be done by IP's or DNS addresses.

\paragraph{Support for Cryptographic Service Provider}
Whether the use, choosing and changing of Cryptographic Service Provider is supported.

@@

\subsubsection{Certificate or certificate chain from a file}
Possibility of extracting certificate or certificate only from a file. Choice of Yes based on possibility of extracting either from a file.

\subsubsection{Private key only}
Possibility of extracting private key only from a certificate.

\subsection{Direct conversion between Java Keystore and PKCS\#12 file}
Possibility of direct conversion (by a command of tested tool) between Java KeyStore and PKCS\#12 type file.

\subsection{Importing of certificates and keys into storage files}
Possibility of importing (additional?) certificates and keys into storage files of said tool. Yes if it is possible to import or add another certificate or key into storage.

@@

\section{General}
Tested under Fedora 21, OpenSSL version 1.0.1k-fips
  \subsection{Type}
  OpenSSL uses PKCS12 to store keys and certificates. Certificates and keys generated in OpenSSL are however being made into PEM or DER encoded file. PEM or DER encoded keys and certificates can then be stored inside single PKCS\#12 file.

@@

  \subsection{License}
OpenSSL is free to use commercialy, however creating CA is not advised in O'REILLY's Network Security with OpenSSL by J. Viega, M. Messier and P. Chandra on page 59,
\begin{displayquote}
Since OpenSSL's command-line CA functionality was intended primarely as an example of how to use OpenSSL to build a CA, we don't recommend that you attempt to use it in a large production environment.
\end{displayquote}
and also hinted in manual pages:
\begin{displayquote}
The ca command is quirky and at times downright unfriendly.

The ca utility was originally meant as an example of how to do things in a CA . It was not supposed to be used as a full blown CA itself: nevertheless some people are using it for this purpose.

The ca command is effectively a single user command: no locking is done on the various files and attempts to run more than one ca command on the same database can have unpredictable results. 
\end{displayquote}

@@

Tested under Fedora 21 and OpenSUSE 42.1, openjdk version 1.8.0\_65

@@

Tested under Fedora 21, OpenSUSE 42.1, Debian Wheezy, gpgsm (GnuPG) 2.0.29, libgcrypt 1.6.3, libksba 1.3.2 and also MS Windows 10 version 1607 Build 14393.447, MS Windows 8.1 Pro 6.3.9600, gpgsm (GnuPG) 2.0.30 (Gpg4win 2.3.3), libgcrypt 1.6.6, libksba 1.3.4

@@

Tested under MS Windows 10 version 1607 Build 14393.447, Keystore Explorer version 5.2.1

@@

Tested under Windows 10 version 1607 Build 14393.447 and Fedora 21, XCA version 1.3.2 using OpenSSL version 1.0.2d

@@

      \paragraph{Self-signed certificate}
You can generate key pair by itself and then use it for generation of certificate. If you do not generate key in advance there is option to generate new key whilst generating a certificate.\textit{{Figure 6.1}} To generate a certificate click on Certificates tab and press New Certificate. There are many options for generating a self-signed certificate, under Source tab you choose a signing algorithm (\verb+MD5|SHA1|SHA224|SHA256|SHA384|SHA512|RIPEMD160+). What must be changed is default hash function because it is set to SHA-1 which is no longer considered safe. Under Subject tab, you shoose Common Name, Organisation, etc. fields of certificate and also choose private key. If you didn't generate key in advance then you can generate a new private key (\verb+RSA|DSA|EC+). For certificate to be accepted by all operating systems you need to choose \verb+Digital Signature & Key Agreement & Certificate Sign+ under Key usage tab. \textit{(Figure 6.2)}
\begin{figure}[!ht]
 \caption{Generation of key for new cert}
 \centering
  \includegraphics[width=0.5\textwidth]{../Dependancies/XCA/genKey.jpg}
\end{figure}

@@

Tested under Fedora 21, certtool version 3.3.16

@@

Tested under MS Windows 10 version 14393.447, makecert.exe version 6.3.9600.17298

@@

Tested under MS Windows 10 version 14393.447

@@

The main thing to understand is, that every tool has had its strong and weak points. Every should be used in complience with required properties of use. That can mean restrictions on used Operating System (e.g. customer has only access to Apple computers with OS X operating system), or for example restrictions how it should be used on Operating System (command line vs GUI programm). In the next part i will try to go over strengths and weaknesses of each tool that I have tested.

\section{OpenSSL}
It is a great command line tool, can work with PEM, DES and PKCS\#12 files. Strong point is a easy form of creating a self-signed certificate, however using it for CA certificates with ca option is not ideal - it is noted in man pages and also a O'REILLY's OpenSSL book.
Another strong point is configuration, which can be done in file. I tested it with creation of appropriate directories and configs.

@@

Main downfall is that keys of length greater than 4096 take longer to generate (notably RSA of length 16192, it took almost 2 hours to generate).

@@

\section{GnuTLS}
Command line tool, availiable only under Linux. Very friendly to use, many different options in configuration file. Great support for different key and hash algorithm types. Supports key lengths of higher values (greater than 4096 RSA for example). Its weaknesses are no support for CSP and availibility only under UNIX like Operating Systems.

@@

>>>>

\section{General}
\begin{table}[h!]
\centering
\caption{Overwiew General}
\label{my-label}
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
                              & \multicolumn{5}{c|}{General}                                                                                   \\ \cline{2-6} 
\multirow{-2}{*}{Tools}       & Type   & View information about certificate                 & License & Operating System                        & Tool Type           \\ \hline
OpenSSL                       & PKCS12 & \cellcolor[HTML]{34FF34}Yes                        & Public  & \cellcolor[HTML]{34FF34}All             & Command Line        \\ \hline
Keytool                       & JKS    & \cellcolor[HTML]{34FF34}{\color[HTML]{000000} Yes} & Public  & \cellcolor[HTML]{34FF34}All             & Command Line        \\ \hline
GnuPG                         & None   & \cellcolor[HTML]{FE0000}No                         & Public  & \cellcolor[HTML]{34FF34}All             & Command Line        \\ \hline
Keystore Explorer             & Both   & \cellcolor[HTML]{34FF34}Yes                        & Public  & \cellcolor[HTML]{FFC702}Win $\vert$ OSX & Graphical Interface \\ \hline
XCA                           & PKCS12 & \cellcolor[HTML]{34FF34}Yes                        & Public  & \cellcolor[HTML]{34FF34}All             & Graphical Interface \\ \hline
GnuTLS                        & PKCS12 & \cellcolor[HTML]{34FF34}Yes                        & Public  & \cellcolor[HTML]{FFC702}UNIX            & Command Line        \\ \hline
Makecert                      & PKCS12 & \cellcolor[HTML]{FE0000}No			    & Public  & \cellcolor[HTML]{FFC702}Windows         & Command Line        \\ \hline
Windows PowerShell PKI Module & PKCS12 & \cellcolor[HTML]{34FF34}Yes			    & Public  & \cellcolor[HTML]{FFC702}Windows         & Command Line        \\ \hline
\end{tabular}
\end{adjustbox}
\end{table}
%TABLE END

\subsection{Type}
This section defines type of storage file in which the certificates or keys are saved. 
Options are:
    \begin{itemize}
     \item \textit{PKCS\#12} - PKCS\#12 encoded file, usual extensions are \verb+.pfx+ and \verb+.p12+
     \item \textit{JKS} - Java KeyStore type file, usual extension is \verb+.jks+
    \end{itemize}

\subsection{View information about certificate}
This section shows whether it is possible to view information certificates and keys and additional information. 
    \begin{itemize}
     \item \textit{Yes} - viewing information about certificates and keys is availiable. This includes length, used algorithm, fingerprint.
     \item \textit{No}  - tool does not have an option to view any information about certificates or keys.
    \end{itemize}

\subsection{License}
Type of license and possibility of using said tool for testing or production code. 
    \begin{itemize}
     \item \textit{Public} - No restrictions for usage, can be used commercialy.
     \item \textit{Testing} - Tool is free to use non-commercialy. For commercial acquisition of paid license is needed.
     \item \textit{Paid} - Every type of usage is paid for. Restrictive version of product may be available, but to use all the features and be able to use in commercial enviroment paid license is needed.
    \end{itemize}

Public does not mean that it should be used as such. Definition if it is advised to use said tool in production code or only in testing enviroment will be talked about in next chapters under each tool.

\subsection{Operating Systems}
In this part the Operating System under which the tool is implemented is defined. 
    \begin{itemize}
     \item \textit{OS X $\vert$ Windows $\vert$ LINUX} - Tool is implemented under any of the mentioned OS, \verb+LINUX+ defines most common distributions of Linux systems \textit{(Debian or RPM based)}.
     \item \textit{UNIX} - Tool is implemented under UNIX-like systems, which is same as \textit{OS X $\vert$ LINUX} option.
     \item \textit{All} - Tool is implemented under every aforementioned operating system.
    \end{itemize}

@@

\paragraph{Self-signed certificate}
Capability of tool to create a self-signed certificate
\begin{itemize}
 \item \textit{Yes} - there is a capability of a tool to create self-signed certificate, not necessarily a single command.
 \item \textit{No} - under no circumstances is it possible to create self-signed certificate using only tool itself.
\end{itemize}

\paragraph{Certificate Signing Request}
Capability of tool to create a Certificate Signing Request.
\begin{itemize}
 \item \textit{Yes} - The tool is capable of creating a certificate signing request, but does not necessarily generates or packs private key with it.
 \item \textit{No} - The tool is not capable of creating any type of certificate signing request.
\end{itemize}

\paragraph{Specify length}
Capability of tool to specify length of created key.
\begin{itemize}
 \item \textit{Yes} - changing key length is possible either in fixed range or by declaring required length \textit{(in bits)}.
 \item \textit{No} - not possible to change key length by any means.
\end{itemize}

\paragraph{Specify algorithm}

\begin{itemize}
 \item \textit{Yes} - changing key or at least hash algorithm is possible .
 \item \textit{No} - not possible to change any algorithm by any means.
\end{itemize}

@@

\paragraph{Specify Type}

\begin{itemize}
 \item \textit{Yes} - Specification if created certificate will be a certificate authority or end certificate is possible.
 \item \textit{No} - Created certificate is of fixed type, either end certificate or certificate authority.
\end{itemize}

\paragraph{Specify Path Length}

\begin{itemize}
 \item \textit{Yes} - Specification of chain length value is available .
 \item \textit{No} - The chain length has a fixed value.
\end{itemize}

@@

\paragraph{Certificate Signing Request signing}

\begin{itemize}
\item \textit{Yes} - The tool supports signing of certificate signing request.
\item \textit{No} - the tool does not support signing of certificate signing requests.
\end{itemize}

\paragraph{Create combination of private key and signed chain}

\begin{itemize}
 \item \textit{Yes} - if it is possible to combine private key and corresponding chain into a single file
 \item \textit{No} - otherwise.
\end{itemize}

\paragraph{Specify certificate validity}

\begin{itemize}
 \item \textit{Yes} - Possibility of choosing how long the certificate will be valid. Done by certificate authority.
 \item \textit{No} - If there is no option to change how long the certificate will be valid, so the duration is a fixed ammount.
\end{itemize}

\paragraph{Setting Subject Alternative Name for end certificates}

\begin{itemize}
 \item \textit{Yes} - possibility of setting a subject alternative name for certificate. It can be DNS name or IP address.
 \item \textit{No} - not possible to choose any subject alternative name.
\end{itemize}

\paragraph{Support for Cryptographic Service Provider}
\begin{itemize}
 \item \textit{Yes} - if there is option to change a cryptographic service provider. 
 \item \textit{No} - the cryptographic service provider canot be changed.
\end{itemize}

@@

\subsubsection{Certificate or certificate chain from a file}
\begin{itemize}
 \item \textit{Yes} - capability of tol to exract certificate or certificate chain from a single file.
 \item \textit{No} - if extracting is not available.
\end{itemize}

\subsubsection{Private key only}

\begin{itemize}
 \item \textit{Yes} - private key can be exported to a single file.
 \item \textit{No} - private key extracting is not allowed.
\end{itemize}

\subsection{Direct conversion between Java Keystore and PKCS\#12 file}

\begin{itemize}
 \item \textit{Yes} - if tool has a capability to convert Java Keystore to PKCS\#12 file and vica versa.
 \item \textit{No} - if it doesn't.
\end{itemize}

\subsection{Importing of certificates and keys into storage files}

\begin{itemize}
 \item \textit{Yes} - it is possible to import certificates and keys into a tools storage type file \textit{(e. g. PKCS\#12 file)}.
 \item \textit{No} - it is not possible to import a certificate into a file.
\end{itemize}

@@

\section{General}
\textit{Tested under Fedora 21, OpenSSL version 1.0.1k-fips}
  \subsection{Type}
  OpenSSL uses PKCS12 to store keys and certificates. Certificates and keys generated in OpenSSL are however being made into PEM or DER encoded file. PEM or DER encoded keys and certificates can then be stored inside single PKCS\#12 file.

@@

  \subsection{License}
OpenSSL is free to use commercialy, however creating CA is not advised in O'REILLY's Network Security with OpenSSL by J. Viega, M. Messier and P. Chandra on page 59,
\begin{displayquote}
\textit{Since OpenSSL's command-line CA functionality was intended primarely as an example of how to use OpenSSL to build a CA, we don't recommend that you attempt to use it in a large production environment.}
\end{displayquote}
and also hinted in manual pages:
\begin{displayquote}
\textit{The ca command is quirky and at times downright unfriendly.}

\textit{The ca utility was originally meant as an example of how to do things in a CA . It was not supposed to be used as a full blown CA itself: nevertheless some people are using it for this purpose.}

\textit{The ca command is effectively a single user command: no locking is done on the various files and attempts to run more than one ca command on the same database can have unpredictable results.}

\end{displayquote}

@@

\textit{Tested under Fedora 21 and OpenSUSE 42.1, openjdk version 1.8.0\_65}

@@

\textit{Tested under Fedora 21, OpenSUSE 42.1, Debian Wheezy, gpgsm (GnuPG) 2.0.29, libgcrypt 1.6.3, libksba 1.3.2 and also MS Windows 10 version 1607 Build 14393.447, MS Windows 8.1 Pro 6.3.9600, gpgsm (GnuPG) 2.0.30 (Gpg4win 2.3.3), libgcrypt 1.6.6, libksba 1.3.4}

@@

\textit{Tested under MS Windows 10 version 1607 Build 14393.447, Keystore Explorer version 5.2.1}

@@

\textit{Tested under Windows 10 version 1607 Build 14393.447 and Fedora 21, XCA version 1.3.2 using OpenSSL version 1.0.2d}

@@

      \paragraph{Self-signed certificate}
You can generate key pair by itself and then use it for generation of certificate. If you do not generate key in advance there is option to generate new key whilst generating a certificate.\textit{{Figure 6.1}}  \\

To generate a certificate click on Certificates tab and press New Certificate. There are many options for generating a self-signed certificate, under Source tab you choose a signing algorithm.  \\

Hash algorithm is one of: \verb+MD5, SHA1, SHA224, SHA256, SHA384, SHA512, RIPEMD160+. It needs to be changed because hash function is set to SHA-1 which is no longer considered safe.  \\

Under Subject tab, you shoose Common Name, Organisation, etc. fields of certificate and also choose private key. If you didn't generate key in advance then you can generate a new private key (\verb+RSA|DSA|EC+). For certificate to be accepted by all operating systems you need to choose \verb+Digital Signature & Key Agreement & Certificate Sign+ under Key usage tab. \textit{(Figure 6.2)}

\begin{figure}[!ht]
 \caption{Generation of key for new cert}
 \centering
  \includegraphics[width=0.5\textwidth]{../Dependancies/XCA/genKey.jpg}
\end{figure}

@@

\textit{Tested under Fedora 21, certtool version 3.3.16}

@@

\textit{Tested under MS Windows 10 version 14393.447, makecert.exe version 6.3.9600.17298}

@@

\textit{Tested under MS Windows 10 version 14393.447}

@@

The main thing to understand is, that every tool has had its strong and weak points. Every should be used in complience with required properties of use.\\

That can mean restrictions on used Operating System \textit{(e.g. customer has only access to Apple computers with OS X operating system)}, or for example restrictions how it should be used on Operating System \textit{(command line vs GUI programm)}. In the next part i will try to go over strengths and weaknesses of each tool that I have tested.

\section{OpenSSL}
It is a great command line tool, can work with PEM, DES and PKCS\#12 files. Strong point is a easy form of creating a self-signed certificate, however using it for CA certificates with ca option is not ideal - it is noted in man pages and also a O'REILLY OpenSSL book.
Another strong point is configuration, which can be done in file. I tested it with creation of appropriate directories and configs.

@@

Main downfall is that keys of length greater than 4096 take longer to generate \textit{(notably RSA of length 16192, it took almost 2 hours to generate)}.

@@

\section{GnuTLS}
Command line tool, availiable only under Linux. Very friendly to use, many different options in configuration file. Great support for different key and hash algorithm types. Supports key lengths of higher values \textit{(greater than 4096 RSA for example)}. Its weaknesses are no support for CSP and availibility only under UNIX like Operating Systems.

@@

\section{Choosing the right tool}
The crucial part of deciding which tool to use is to know what is desired. The first choice to make is to decide what file type will be required.
In this list there are most useful tools for each type:

\begin{itemize}
 \item \textbf{PKCS\#12} \begin{enumerate}\item \textit{GUI} - XCA \item \textit{Command Line} - OpenSSL \end{enumerate}
 \item \textbf{Java KeyStore} \begin{enumerate}\item \textit{GUI} - Keystore Explorer \item \textit{Command Line} - keytool \end{enumerate}
\end{itemize}

Other tested tools can be also used, but \textit{OpenSSL, Keytool, XCA} and \textit{Keystore Explorer} can do almost everything, along with their ease of use. My personal favourite has became \textit{XCA} as it is tool where i checked if information about certificates from other tools checked out.\\

But only because they can do everything doesn't mean they are best at everything, for example if self-signed certificate is neede only for testing purposes, one can be really quickkly made with help Windows PowerShell PKI Module where no additional instalations are required.\\

Considering Java KeyStore tools, only 2 were tested, 1 GUI and 1 Command Line. Nevertheless, even though they have their own upsides and downsides, are easy and quickly ready to use.\\
